---
title: Classes and Methods
subtitle: R Programming
author: 
  - name: Ella Kaye and Heather Turner
  - name: Department of Statistics, University of Warwick
title-slide-attributes:
  data-background-color: "#552D62"
date: 2024-03-18
date-format: long
format: 
  warwickpres-revealjs:
    execute:
      echo: true
      code-overflow: wrap
---

```{r}
#| label: setup
#| include: false
options(digits = 4)
```


## Overview

- Object-oriented programming
- S3
- Other OOP systems (S4, R6, S7)

## Source material

This material is largely based on Chapters 12 and 13 of Advanced R, 2nd edition, by Hadley Wickham.

The book is freely available online: <https://adv-r.hadley.nz>.

It is shared under the [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/).

# Object-oriented programming {.inverse}

## Object-oriented programming (OOP)

With OOP, a developer can consider a function's interface separately from its implementation. 

This makes it possible to use the same function for different types of input. 

These are called **generic** functions. 

## OOP definitions

- OO systems call the type of an object its **class**.

- An implementation for a specific class is called a **method**.

- The class defines the **fields**, the data possessed by every instance of that class.

. . .

:::{.callout-tip}
Roughly speaking, a class defines what an object *is* and methods define what an object can *do*.
:::

## OOP definitions (continued)

- Classes are organised in a hierarchy, so that if a method does not exist for one class, its parent's method is used.

- The child is said to **inherit** behaviour.

- The process of finding the correct method given a class is called **method dispatch**.

## Generic functions

**Generic** functions provide a unified interface to **methods** for objects of a particular **class**, e.g.

```{r penguins}
library(palmerpenguins)
```

```{r generic}
summary(penguins$species)
summary(penguins$flipper_length_mm)
```

Here, we use the same function, `summary()`, on objects of classes `factor` and `integer` and get different output for each.

## Motivation for OOP

`summary()` could contain several if-else statements, but 
 
 - the code would become hard to follow
 - only the function authors (R Core) could add new implementations
 
## What does OOP offer?

 - separates the code for different data types
 - avoids duplicating code by method inheritance from parent class(es) to child class (subclass)
 - makes it possible for external developers to add methods for new types of object
    - this can be particularly useful when writing R packages


## OOP Systems in R

There are 3 main OOP systems in use:

:::{.smaller90}
* **S3** 
  - Used in base R and most recommended/CRAN packages
  - Use unless you have good reason not to
* **S4**
  - Used on Bioconductor
  - Allow more complex relationships between classes and methods
* **R6**
  - More similar to OOP in other languages
  - May prefer if S3 insufficient and not aiming for Bioconductor
    
A new OOP system, **S7**, is in [development](https://rconsortium.github.io/OOP-WG/) as a successor to **S3** and **S4**. 
:::

:::{.notes}
R6 v similar to Python OOP (also Java, C++)
Julia not an OOP language, but does have concept of multiple dispatch
- so maybe a bit more like S3 idea?
:::

## sloop

The **sloop** package provides tools to help you interactively explore and understand object oriented programming in R, particularly with S3.

```{r}
library(sloop)
```

## Objects

In R, we can distinguish between **base objects** and **OO objects**.

A base object:
```{r}
is.object(1:10)
sloop::otype(1:10)
```

An OO object:
```{r}
is.object(penguins)
sloop::otype(penguins)
```


::: {.notes}
Everything in R is an object, but not everything is object-oriented.  
:::

## Classes

Techincally, the difference between base and OO objects is that OO objects have a **class** attribute:

```{r}
attr(1:10, "class")
attr(penguins, "class")
sloop::s3_class(penguins)
```

## Base types

Only OO objects have a class attribute, but every object has a **base type**.

There are 25 different base types, e.g.

```{r}
typeof(NULL)
typeof(1)
typeof(1L)
typeof("hello")
```
## "object of type 'closure' is not subsettable"

```{r error = TRUE}
typeof(mean)
mean[1]
```

. . . 

Jenny Bryan's talk on debugging: 

<https://posit.co/resources/videos/object-of-type-closure-is-not-subsettable/>

# S3 {.inverse}

## S3 objects

An S3 object has a `"class"` attribute:

```{r}
attr(penguins$species, "class")
unique(penguins$species)
```

## S3 objects: the underlying object

With `unclass()` we obtain the underlying object, its **base type**, here an integer vector 

```{r}
species_no_class <- unclass(penguins$species)
class(species_no_class)
unique(species_no_class)
attributes(species_no_class)
```
## OO type vs base type when passed to generic

```{r}
f <- factor(c("a", "b", "c"))
print(f)
print(unclass(f))
```

## generic as middleman

> The generic is the middleman: its job is to define the interface (i.e. the arguments) then find the right implementation for the job. The implementation for a specific class is called a **method**, and the generic finds that method by performing **method dispatch**.

Hadley Wickham, Advanced R (2e)

## Naming scheme

S3 methods are functions with a special naming scheme, `generic.class()`. For example, the `factor` method for the `print()` generic is called `print.factor()`.

You should never call the method directly, but instead rely on the generic to find it for you.

. . . 

:::{.callout-tip}
This is why it is not considered best practice to use `.` when naming your own functions.
:::

. . .

:::{.callout-warning}
Lots of important R functions that are **not** methods *do* have `.` in the title -- these are from before S3.
:::

::: {.notes}
e.g.
data.frame()
row.names()
:::

## Creating OO objects

To make an object an instance of a class, you simply set the **class attribute**.

(S3 has no formal definition of a class).

## Creating an S3 object with `stucture()`

You can use `structure()` to define an S3 object with a class attribute:

```{r}
dp <- 2
structure(list(pi = trunc(10^dp * pi)/10^dp, dp = dp),
          class = "pi_trunc")
```

Potentially further attributes can be added at the same time, but typically we would use a list to return all the required values.

## Creating an S3 object with `class()`

Alternatively, we can add a class attribute using the `class()` helper function:

```{r}
pi2dp <- list(pi = trunc(10^dp * pi)/10^dp, dp = dp)
class(pi2dp) <- "pi_trunc"
pi2dp
```

## Warning!

S3 has no checks for correctness, so we can change the class of objects. 

**This is a bad idea!**

```{r error = TRUE}
mod <- lm(flipper_length_mm ~ bill_length_mm, data = penguins)
class(mod)
class(mod) <- "Date"
print(mod)
```

. . .

> R doesn't stop you from shooting yourself in the foot, but as long as you don't aim the gun at your toes and pull the trigger, you won't have a problem.

## Creating your own classes

All objects of the same class should have the same structure, i.e. same base type and same attributes.

Recommend that you create:

- a low-level **constructor**, `new_myclass()`, that efficiently creates objects with the correct structure
- A **validator**, `validate_myclass()` that performs more computationally expensive checks to ensure the object has correct values
- A user-friendly **helper**, `myclass()`, that provides a convenient way for others to create objects of your class.

See <https://adv-r.hadley.nz/s3.html#s3-classes> for more details.

## print generic

```{r}
print(pi2dp)
```

:::{.notes}
See what happens before we define a custom print method later

Prints like a list along with the attribute
:::

## S3 generic functions

S3 generic functions are simple wrappers to `UseMethod()`

```{r}
print
```

## `useMethod()`

The `UseMethod()` function takes care of *method dispatch*: selecting the S3 method according to the class of the object passed as the first argument.


```{r}
class(penguins$species[1:3])
print(penguins$species[1:3])
```

Here `print()` dispatches to the method `print.factor()`. 

## `s3_dispatch()`

`UseMethod()` creates a vector of method names then looks for each potential method in turn. We can see this with `sloop::s3_dispatch()`:

```{r}
s3_dispatch(print(penguins$species))
```

- `=>` indicates the method that is called here.
- `*` indicated a method that is defined, but not called.

## `default`

`default` is a special **pseudo-class** that provides a fallback whenever a class-specific method is not available.

```{r}
s3_dispatch(print(pi2dp))
```

`print.pi_trunc` is not defined.

## Method dispatch

An S3 object can have more than one class e.g.
```{r}
class(penguins)
```
`UseMethod()` works along the vector of classes (from the first class to the last), looks for a method for each class and dispatches to the first method it finds.

If no methods are defined for any of class, the default is used , e.g. `print.default()`.

If there is no default, an error is thrown.

## S3 methods for a class

See the methods for a given S3 class:

```{r}
# nls is nonlinear least squares
methods(class = "nls")
```
. . . 

```{r}
s3_methods_class("nls") |> head()
```

## S3 methods for a generic

See the methods for a given generic function:

```{r}
methods("coef")
```
Asterisked methods are not exported.

. . .

```{r}
s3_methods_generic("coef")
```

## View S3 methods

S3 methods need not be in the same package as the generic.

Find an unexported method with `getS3method()` or `sloop::s3_get_method()`

```{r}
getS3method("coef", "default")
```
```{r}
#| eval: false
s3_get_method("coef.default") # equivalent
```


::: {.notes}
Can view code for exported functions by just typing the function name, or, in RStudio, using `View(function_name)`  
:::

## Writing S3 Methods

The arguments of a new method should be a superset of the arguments of the generic

```{r}
args(print)
```

. . .

New methods have the name format `generic.class`:

```{r}
print.pi_trunc <- function(x, abbreviate = TRUE, ...){
  dp_text <- ifelse(abbreviate, "d.p.", "decimal places")
  cat("pi: ", x$pi, " (", x$dp, " ", dp_text, ")", sep = "")
}
print(pi2dp)
```

. . . 

```{r}
print(pi2dp, abbreviate = FALSE)
```

::: {.notes}
Note the difference compared to calling `print(pi2dp)` before.
:::

## Inheritance

S3 classes can share behaviour through a mechanism called **inheritance**. Inheritance is powered by three ideas.

- The class can be a character *vector*

- If a method is not found for the class in the first element of the vector, R looks for a method in the second class (and so on)

- A method can delegate work by calling `NextMethod()`.

## Multiple classes

The class of an S3 object can be a vector of classes: 

```{r}
fit <- glm(y ~ x, data = data.frame(y = 1:3, x = 4:6))
class(fit)
```
We say `fit` is a `"glm"` object that **inherits** from class `"lm"`.

- `glm` is a **subclass** of `lm`, because it always appears before it in the class vector.

- `lm` is a **superclass** of `glm`.

::: {.notes}
Good practice:

- base type of the subclass should be the same as the superclass

- attributes of the subclass should be a superset of the attributes of the superclass.

These practices are not enforced. These are recommendations in Advanced R for make life easier.  
:::

## `inherits()`

The `inherits()` function can be used to test if an object inherits from a given class:

```{r}
inherits(fit, "glm")
inherits(fit, "lm")
inherits(fit, "xlm")
```


## Your turn (part 1)

```{r}
#| echo: false
ols <- function(x, y, ...){
  fit <- lm(y ~ x, data =  data.frame(x = x, y = y))
  class(fit) <- c("ols", class(fit))
  fit
}

print.ols <- function(x, ...){
  cat("Intercept: ", coef(x)[1], "\n")
  cat("Slope: ", coef(x)[2], "\n")
}
```

1. Create a function to fit an ordinary least squares model given a response `y` and an explanatory variable `x`, that returns an object of a new class `"ols"`, that inherits from `"lm"`.

2. Define a print method for your function that it works as follows:

```{r}
set.seed(1)
res <- ols(x = 1:3, y = rnorm(3))
res
```

Note: I have set `options(digits = 4)` to limit the number of digits printed by default throughout this presentation (default is 7).


## `NextMethod()`

Hard to understand, so here's a concrete example for the common use case: `[`.

```{r}
new_secret <- function(x = double()) {
  stopifnot(is.double(x))
  structure(x, class = "secret")
}

print.secret <- function(x, ...) {
  print(strrep("x", nchar(x)))
  invisible(x)
}

x <- new_secret(c(15, 1, 456))
x
```

::: {.notes}
?`[`
:::

## But there's a problem

```{r}
x[1]
```

We want this to be secret!
. . . 

The default `[` method doesn't preserve the class.

```{r}
s3_dispatch(x[1])
```

::: {.notes}
There's no `[.default` defined (no `*` next to it), just an internal generic. We'll learn more about those later.  
:::

## A first attempt at a solution

So, need to defined a `[.secret` method. 

But the following doesn't work:

```{r error = TRUE}
`[.secret` <- function(x, i) {
  new_secret(x[i])
}
x[1]
```

It gets stuck in infinite loop.


::: {.notes}
Trying to subset `x` then call new_secret on it, to give it the `secret` class, so that `print.secret` works on said subset.  
:::

## What's the fix?

We need some way of calling the underlying `[` code, i.e. the implementation that would get called if we didn't have a `[.secret` method.

```{r}
`[.secret` <- function(x, i) {
  new_secret(NextMethod())
}
x[1]
```

. . . 

i.e. we're defining `[.secret` but we still want to access the internal `[` method (so we don't get stuck in a loop) as if `[.secret` wasn't defined.


## Delegation with `NextMethod()`

```{r}
s3_dispatch(x[1])
```

The `=>` indicates that `[.secret` is called, but that `NextMethod()` delegates work to the underlying internal `[` method, as shown by `->`.


## Another `NextMethod()` example

```{r}
data <- data.frame(x = 1:3, y = 4:6)
class(data)
data
t(data)
```

::: {.notes}
`t()` is the transpose function. 
We started with a data.frame but after calling `t` on it, we have a matrix. What happened?
:::

## Underlying code

```{r}
t.data.frame
```
```{r}
s3_dispatch(t(data))
```

. . .

We can explicitly call the next method that would be called by `UseMethod()` to reuse code whilst customising as required.

:::{.notes}
We have seen that `UseMethod()` will dispatch to the method of the next class in the class vector if a method is not defined for the first class.
:::

## Implicit classes

As we've seen, `is.object()` or `sloop::otype()` can be used to find out if an object has a class (S3/S4/R6)

```{r}
is.object(factor(1:3))
is.object(1:3)
```

An object that does not have an explicit class has an implicit class that will be used for S3 method dispatch. 

## Implicit classes and dispatch

The implicit class can be found with `.class2()`, or `sloop::s3_class()`

```{r}
M <- matrix(1:12, nrow = 4)
attr(M, "class")
.class2(M)
s3_class(M)
```

The `class()` of an object does not uniquely determine its dispatch:

```{r}
s3_dispatch(print(M))
```

:::{.notes}
.class2 since R 4.0.0. 
:::

## Attributes
:::{.smaller90}
We can take advantage of existing S3 methods by returning an object of a existing S3 class or an implicit class, using attributes to add custom information

```{r}
x <- matrix(c(1:5, 2*(1:5)), ncol = 2)
center_x <- scale(x, scale = FALSE)
class(center_x)
summary(center_x)
```

```{r}
attr(center_x, "scaled:center")
```
This can avoid the need to define new classes and methods, in simple cases.
:::

::: {.notes}
`attr(x, "scaled:center")` in `NULL`

`center_x` is still a matrix, and so `summary(center_x)` dispatches to `summary.matrix`. We can take advantage of the methods for `matrix`, without needing to define a new class for `scaled`.
:::

## Under the hood

```{r}
s3_dispatch(scale(x, scale = FALSE))
```
```{r}
s3_dispatch(summary(center_x))
```
```{r eval = FALSE}
View(scale.default)
```

. . .

In `scale.default()` the attribute `"scaled:center"` is added to the `x` argument, so essentially, `center_x` is a matrix with extra information (in this case, the colMeans of the original columns).

## Your turn (part 2)

```{r}
#| echo: false
summary.ols <- function(object, ...){
  res <- NextMethod()
  class(res) <- c("summary.ols", class(res))
  res
}
print.summary.ols <- function(x, ...){
  cf <- coef(x)
  cat("Coefficients: \n") 
  print(cf)
  cat("\n")
  cat("Residual standard error: ", x$sigma, "\n")
  cat("Multiple R-squared: ", x$r.squared, "\n")
}
```

3. Write a summary method for your `ols` class that uses `NextMethod()` to compute the usual `lm` summary, but return an object of class `"summary.ols"`. 

4. Write a print method for the `"summary.ols"` which works as follows:

```{r}
summary(res)
```

:::{.notes}
will get issues with inheriting e.g. add1 method
:::

# Other OOP systems {.inverse}

## S4

S4 provides a formal approach to OOP. Its implementation is much stricter than S3.

S4 has **slots**, a named component of the object accessed with `@`.

S4 methods:

 - use specialised functions for creating classes, generics and methods
 - allow _multiple inheritance_: a class can have multiple parents
 - allow _multiple dispatch_: method selection based on the class of multiple objects

:::{.notes}
use import instead

The **methods** package provides the functions required to use S4 classes and methods, so always load this package when using S4.
:::

## S4 uses

- S4 is the OOP system used for bioconductor packages
- The **Matrix** package

## R6

- The R6 OOP system is defined in the **R6** package: <https://r6.r-lib.org>
- Encapsulated OOP, similar to OOP systems in other languages
- The Advanced R book cautions against using R6 - it leads to non-idiomatic R code.

## S7

- Currently in the **S7** package: <https://rconsortium.github.io/S7/>
- Will eventually be incorporated into R as the successor S3 and S4. 
- An example of using S7 for generative art: <https://blog.djnavarro.net/posts/2024-02-25_s7/>

# End matter {.inverse}

## References

- Main reference for this session, goes a bit further (including R6):
Wickham, H, _Advanced R_ (2nd edn), _Object-oriented programming section_, <https://adv-r.hadley.nz/oo.html>

- Fun example creating `Turtle` and `TurtleWithPen` classes to create simple graphics by moving the turtle:
<https://stuartlee.org/2019/07/09/s4-short-guide/> 

## License

Licensed under a Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License ([CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/){target="_blank"}).
